---
title: "Causal Effect Estimation on Genotype Data: GLM, TMLE, and C-TMLE"
subtitle: "Causality in Biomedicine: Going Beyond Associations (EMBL-EBI, 4-9 Oct 2026)"
author: "Asantewaa Sarpong & Sjoerd Beentjes"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Purpose

In this notebook we estimate the **average treatment effect (ATE)** of a
single genetic variant (SNP) on a simulated outcome, using three
increasingly sophisticated estimators:

1.  **GLM-based ATE**: a plain regression adjustment. Fast, simple, but
    relies on correctly specifying the outcome model.
2.  **TMLE** (Targeted Maximum Likelihood Estimation): a doubly-robust,
    semiparametric estimator that combines flexible outcome and
    treatment ("propensity") models.
3.  **C-TMLE** (Collaborative TMLE): builds the propensity model
    *collaboratively* with the outcome model, rather than fitting it
    independently, which can help when candidate confounders are highly
    correlated (as genetic variants often are, due to linkage
    disequilibrium).

TMLE and C-TMLE here are **this project's own implementations**
(`src/estimators.R`, `src/utils.R`) , the same code used in the full
simulation benchmark behind this repository, not the bare CRAN package
functions. This matters: `estimators.R` builds in cross-fitting and a
bootstrap-based standard error validated specifically for this genotype
setting, and it defines its own function literally called `tmle()`,
which is a different (and the one we actually want) implementation from
the `tmle` package's own `tmle()`.

We use the `data/geno.csv` genotype matrix from this repository, which
is **fully synthetic**, no real individuals are involved.

Because the true treatment effect is known by construction in this
exercise (we define it ourselves, below), you can directly compare each
estimator's output against the right answer.

```{r source-project-code}
# Sourcing estimators.R also sources utils.R (see its own source() call at
# the top of the file) and loads every package both files need internally
# (glmnet, glm2, caret, tibble, dplyr, tidyr, ggplot2, ctmle) — no separate
# library() calls needed here.
source("src/estimators.R")
```

## 1. Load the data and pick a treatment SNP

```{r load-data}
geno <- read.csv("data/geno.csv", row.names = 1, check.names = FALSE)
geno <- as.matrix(geno)

dim(geno)   # samples x SNPs
```

For this walkthrough we'll treat one SNP as the "treatment" and a small
set of nearby SNPs as potential confounders. In the full SIM2 study (see
the project's `data/` and `src/` folders) every one of 9 treatment SNPs
is tested against \~199 confounders across 200 replicates , here we use
far fewer of both, purely so each estimator fits in a few seconds during
the session.

```{r pick-snp}
treatment_snp    <- colnames(geno)[1]        # first SNP in the matrix
confounder_snps  <- colnames(geno)[2:21]     # next 20 SNPs as confounders

# Binary treatment: carries >=1 copy of the variant vs. none
A <- as.numeric(geno[, treatment_snp] > 0)

# IMPORTANT: this project's estimator functions (tmle(), ctmle1(), ...)
# all expect confounder columns named W1, W2, ... — .W_matrix() in
# utils.R picks them out with a "^W" regex. Using the raw SNP rsIDs as
# column names here would cause .W_matrix() to silently find ZERO
# confounder columns.
W <- as.data.frame(geno[, confounder_snps])
colnames(W) <- paste0("W", seq_len(ncol(W)))

table(A)
```

## 2. Simulate a known-truth outcome

To actually demonstrate GLM's limitations (not just assert them), the
outcome needs to be something a plain `Y ~ A + W` GLM genuinely cannot
represent, not just "nonlinear in W", which turns out to barely bias GLM
at all (see note below). The reliable mechanism is **effect
modification**: we let the *treatment effect itself* vary with `W4`, a
confounder in strong LD with (i.e. correlated with) the treatment SNP.

The population-average (marginal) ATE is still exactly **1** by
construction, we center `W4` so its mean contribution is zero, but no
single additive coefficient on `A` can capture an effect that genuinely
differs by subgroup. If you've taken a GLM course, this is the single
most important failure mode to know: an omitted `A:W` interaction biases
the ATE whenever the modifying covariate correlates with treatment,
which is exactly the situation genetic confounders are usually in.

```{r simulate-outcome}
set.seed(42)

true_ate <- 1
effect_modifier_strength <- 0.5   # how much the treatment effect varies with W4

# Individual treatment effect varies with W4. Centering W4 keeps the
# population-average (marginal) ATE at exactly `true_ate`.
W4_centered <- W$W4 - mean(W$W4)
individual_effect <- true_ate + effect_modifier_strength * W4_centered

Y <- individual_effect * A +
  0.5 * W$W1 - 0.3 * W$W2 + 0.2 * W$W3 +
  rnorm(nrow(W), mean = 0, sd = 1)
```

> **Note on an earlier version of this DGM:** we originally tried adding
> interactions among confounders (`W x W`, no treatment involved) to
> break GLM's linear specification. That turns out not to reliably bias
> GLM's point estimate, once main effects are controlled for, a
> properly-scaled `W x W` interaction term ends up nearly uncorrelated
> with treatment, so GLM's omitted-variable bias from it is small (this
> was a real, hard-won finding from the full simulation study behind
> this repo). Effect modification (`A x W`) is the more reliable demo
> because the omitted term directly involves treatment itself.

Every estimator below is given exactly the same input , a single data
frame with columns `A`, `W1`...`W20`, and `Y` , matching the format this
project's pipeline uses throughout (see `ctmle_mc_analysis.R`).

```{r assemble-data}
data <- data.frame(A = A, W, Y = Y)
```

## 3. GLM-based ATE

The simplest approach: regress the outcome on treatment and confounders,
and read off the coefficient on `A`. There's no project-specific GLM
function for this , it's intentionally just base R's `glm()` , so this
is the one estimator here that is *not* from `src/estimators.R`.

```{r glm-ate}
glm_fit <- glm(Y ~ ., data = data, family = "gaussian")

glm_ate <- coef(glm_fit)["A"]
glm_se  <- summary(glm_fit)$coefficients["A", "Std. Error"]

cat(sprintf("GLM ATE estimate: %.3f (SE: %.3f)\n", glm_ate, glm_se))
```

**Discussion point:** `Y ~ .` fits one constant coefficient on `A`, it
cannot represent a treatment effect that varies with `W4`, the way we
just built into the data. Because `W4` correlates with treatment (strong
LD with the treatment SNP), this constant-effect assumption produces
real bias, not just extra noise. And critically: nothing about the GLM
output above tells you this happened, the model fits "successfully" and
reports a confident standard error regardless.

## 4. TMLE

TMLE improves on the GLM approach in two ways: it allows flexible,
cross-fitted models for both the outcome and the propensity score, and
it "targets" the outcome model to the parameter of interest (the ATE),
giving an estimator that is **doubly robust** , consistent if *either*
the outcome model or the propensity model is correctly specified, not
necessarily both.

We call this project's `tmle()` directly, with its production defaults:
cross-fitted ("CV") variance estimation, and a targeted bootstrap for
the standard error (`n_boot = 200` resamples by default , if this feels
slow on your machine during the session, try `n_boot = 50`).

> **Important, and worth saying out loud to participants:** this
> project's `tmle()`/`ctmle1()` build their own outcome model the same
> way GLM does, main effects only, no `A:W4` interaction (see
> `.make_QX()` in `utils.R`). So they are not automatically immune to
> the effect-modification bias above. Their protection, if any, comes
> entirely from TMLE's **doubly-robust** property: consistent if
> *either* the outcome model or the propensity model is correct, not
> necessarily both. Whether that actually rescues the estimate here is
> an empirical question, look at the bias numbers in Section 6, don't
> assume the answer.

```{r tmle-fit}
tmle_fit <- tmle(data, alpha = 0.05)

tmle_ate <- tmle_fit$est
tmle_se  <- tmle_fit$se

cat(sprintf("TMLE ATE estimate: %.3f (95%% CI: %.3f, %.3f)\n",
            tmle_ate, tmle_fit$ci_lower, tmle_fit$ci_upper))
```

## 5. C-TMLE

C-TMLE is most useful when candidate confounders are highly correlated
with each other or with treatment , exactly the situation genetic
variants are often in, due to linkage disequilibrium. Rather than
fitting the propensity model independently of the outcome model, C-TMLE
builds it up collaboratively: here, via a LASSO path over the propensity
model with the penalty strength chosen by targeted cross-validated loss,
rather than ordinary predictive CV.

We use `ctmle1()`, this project's primary C-TMLE variant (there are
several others in `src/estimators.R` , `ctmle0()`, `scalable_ctmle()`,
the Wang-CTMLE family , each making a different collaborative-selection
choice; `ctmle1()` is the one used as the main comparison point
throughout the SIM2 study).

```{r ctmle-fit}
ctmle_fit <- ctmle1(data, alpha = 0.05)

ctmle_ate <- ctmle_fit$est
ctmle_se  <- ctmle_fit$se

cat(sprintf("C-TMLE ATE estimate: %.3f (95%% CI: %.3f, %.3f)\n",
            ctmle_ate, ctmle_fit$ci_lower, ctmle_fit$ci_upper))
```

## 6. Compare all three

```{r compare}
results <- data.frame(
  Method   = c("GLM", "TMLE", "C-TMLE"),
  Estimate = c(glm_ate, tmle_ate, ctmle_ate),
  Bias     = c(glm_ate, tmle_ate, ctmle_ate) - true_ate
)

print(results, row.names = FALSE)
```

**Discussion questions for participants:**

- How close is each estimate to the true ATE of 1?
- Which estimator's confidence interval actually contains the true
  value?
- What happens if you pick a *different* treatment SNP , particularly
  one more strongly correlated with its neighbouring confounders? (Try
  changing `treatment_snp` above to a SNP from later in the matrix and
  re-running.)
- The full simulation study behind this repository runs this same
  comparison across 9 treatment SNPs spanning Strong/Moderate/Minimal
  linkage disequilibrium, with 200 replicates each, and finds that
  **confidence interval coverage degrades under strong LD** even when
  point estimates look reasonable. Why might that be, given what you've
  just seen GLM, TMLE, and C-TMLE each do differently?

## Session info

```{r session-info}
sessionInfo()
```
